Distributional Trees for Circular Data

### 2019-04-08 pkg/disttree/inst/IWSM2019/abstract_IWSM2019_schlosser.tex
further revisions on extended abstract
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@2520 255f4249-2263-4417-b1ea-8498779e27a7

Former-commit-id: 96a3aaedbf6b3c4a4f08c16cc4e3c29aead080ab

%\section{Motivation}
%Circular response variables occur in a variety of applications and subject areas. 
%For example, gun crime data on a $24$-hour scale is analysed in the social-economics, 
%animal orientation or gene-structure analysis are often subject of examination in biology, 
%and wind direction is one of the most important weather variables in meteorology.
%Representing such a two-dimensional direction by an angle in a unity circle requires 
%the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
%rotation. However, these settings can vary but should not influence statistical 
%inference from a fitted model.
%For that reason circular data demand for purpose-built statistical methods. 
%A commonly chosen distribution for modeling circular data is the von Mises distribution.
%Estimating its two distribution parameters by generalized additive models using a set of 
%regressor variables is one possible approach that can be successful in various settings.
%As an alternative we propose the application of distributional trees 
%which are capable of detecting also non-additive effects and provide easily interpretable 
%models allowing for a wide range of inference procedures.
%Moreover, covariates and their possible interactions are selected automatically
%such that no expert knowledge is required to specify the model in advance.

@@

A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model
(see Schlosser et.~al~(2018) for a general discussion of distributional forests).

Another possible improvement for obtaining more parsimonious and more stable trees is to
consider splitting circular covariates into two circle segments by searching two
split points simultaneously (rather than sequentially at different depths as
for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).

\textsf{R} packages implementing the proposed methods are under development at
\texttt{https://R-Forge.R-project.org/projects/partykit/}.

@@

@@

%\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
%     Classification and Regression Trees.
%     {\it Wadsworth}.

@@

>>>>

\section{Motivation}
%\textit{FIXME: should we include this section 'Motivation'?}

Circular data can be found in a variety of applications and subject areas, 
e.g., hourly crime rate in the social-economics, 
animal movement direction or gene-structure in biology, 
and wind direction as one of the most important weather variables in meteorology.
Circular regression models were first introduced by Fisher and Lee (1992) and 
further extended by Jammalamadaka and Sengupta (2001) and 
Mulder and Klugkist (2017) among others.
While most of these approaches are built on additive regression models, 
we propose an adaption/extension of regression trees to circular data by employing 
distributional trees.

%Representing such a two-dimensional direction by an angle in a unity circle requires 
%the specification of a starting point (angle 0) and either a clockwise or anti-clockwise 
%rotation. However, these settings can vary but should not influence statistical 
%inference from a fitted model.
%For that reason circular data demand for purpose-built statistical methods. 
%A commonly chosen distribution for modeling circular data is the von Mises distribution.
%Estimating its two distribution parameters by generalized additive models using a set of 
%regressor variables is one possible approach that can be successful in various settings.
%As an alternative we propose the application of distributional trees 
%which are capable of detecting also non-additive effects and provide easily interpretable 
%models allowing for a wide range of inference procedures.
%Moreover, covariates and their possible interactions are selected automatically
%such that no expert knowledge is required to specify the model in advance.

@@

\subsection{Ensembles/Random Forests}
A natural extension are ensembles (or forests) of such circular trees
that can improve forecasts by regularizing and stabilizing the model.
Random forests introduced by Breiman~et.~al~(1984) average the
predictions of an ensemble of trees, each built on a subsample 
or bootstrap of the original data. 
A different strategy of combining trees is employed in distributional forests 
(Schlosser~et.~al, 2019) which can also be seen as adaptive local likelihood models. 
For a possibly new observation $x$ weights $w_i(x)$ are obtained by counting the 
number of trees in which it is assigned to the same terminal node as a learning
observation $y_i, i \in \{1,\ldots,n\}$.
The parameters $\mu$ and $\kappa$ are then estimated specifically for the (new) 
observation $x$ by applying the weighted maximum likelihood estimator
$$
\operatorname{argmax}\displaylimits_{\mu, \kappa} \sum_{i=1}^n w_i(x) \cdot \ell(\mu, \kappa; y_i). 
$$
In that way predictions are mainly based on similar observations rather than the 
complete learning data.

\subsection{Splits in Circular Variables}
Another planned extension of the introduced circular distributional trees 
for obtaining more parsimonious and more stable trees is to
consider splitting circular covariates into two flexible circle segments by searching two split 
points simultaneously.
In general, the search for the best separation of the data space into two intervals is based on the optimization problem
$$
\max \left(\sum_{y \in \mathit{left}} \ell(\hat{\mu_1}, \hat{\kappa_1}; y) + \sum_{y \in \mathit{right}} \ell(\hat{\mu_2}, \hat{\kappa_2}; y)\right)
$$
where $\hat{\mu_1}$, $\hat{\kappa_1}$, $\hat{\mu_2}$, $\hat{\kappa_2}$ are the estimated parameters for the corresponding subsets of the response as defined by the intervals, here generally denoted by \textit{``left''} and \textit{``right''}.
For a linear split variable these intervals are defined by a single split point.
By treating a circular variable such as wind direction linearly
this single split point $\nu$ devides the considered space~$[0,2\pi)$ into
$\mathit{left}=[0,\nu]$ and $\mathit{right}=(\nu,2\pi)$ enforcing a potentially unnatural 
separation at zero. 
Moreover, further sequential splits in the circular variable are often required to obtain 
data driven subintervals of~$[0,2\pi)$ (as seen for the lagged wind direction in Figure~\ref{schlosser:fig_tree}).
This can be avoided by searching for two split points $\nu$ and $\tau$ within the maximization
process such that $\mathit{left}=[\nu,\tau]$ and $\mathit{right}=[0,\nu) \cup (\tau,2\pi]$
where both interval borders are determined purely by the data.

The required test statistic to maximally select two split points simultaneously is readily 
available in the \textsf{R} package \texttt{coin} but would need to be further enhanced as proposed 
by Hothorn and Zeileis~(2008). For a score-based approach the general strategy of double maximum tests introduced by Bai and Perron~(2003) can be considered. 
Hence, the splitting idea can be naturally extended to a two-point search, however, for 
an unbiased and inference based selection the corresponding tests employing a proper
asymptotic null distribution still need to be implemented.

@@

\subsubsection*{Computational Details:}
\textsf{R} packages implementing the proposed methods are currently under development at
\texttt{https://R-Forge.R-project.org/} \texttt{projects/partykit/}.

@@

\item[Bai, J., and Perron, P.] (2003).
     Computation and Analysis of Multiple Structural Change Models.
     {\it Journal of Applied Econometrics}, {\bf 18},
     1\,--\,22.

\item[Breiman, L., Friedman, J. H., Olshen, R. A., and Stone, C. J.] (1984).
     Classification and Regression Trees.
     {\it Wadsworth}.

\item[Fisher, N. I., and Lee, A. J.] (1992).
     Regression Models for an Angular Response.
     {\it Biometrics}, {\bf 48}, 3,
     665\,--\,677.

@@

\item[Mulder, K., and Klugkist, I.] (2017).
     Bayesian Estimation and Hypothesis Tests for a Circular Generalized Linear Model.
     {\it Journal of Mathematical Psychology}, {\bf 80},
     4\,--\,14.
